---
title: |
  | Rachel Hardy
  |
  | ST 563 (601)
  |
  | Final Project
  |
  | Due 12/10/24
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tinytex)
library(tidyverse)
library(readr)
library(dplyr)
library(knitr)
library(caret)
library(rsample)
library(MASS)
library(ISLR2)
library(corrplot)
library(pls)
library(leaps)
library(glmnet)
library(naivebayes)
library(gam)
library(splines)
library(boot)
library(ggplot2)
library(GGally)
library(data.table)
library(earth) # For MARS models
library(rpart) # For regression trees
library(rpart.plot) # For plotting regression trees
library(randomForest) # For random forests
library(e1071) # For calculating variable importance
library(ipred) # For fitting bagged decision trees
library(BART) # For Bayesian Additive Regression Trees
library(gbm) # For boosted regression trees
library(rJava)
library(bartMachine) # For Bayesian Additive Regression Trees
library(h2o)
```

\newpage

# Data Preparation

**Below we will prepare the data!**

```{r}
## Original Data ----
bleaching_data <- fread("global_bleaching_environmental.csv", 
                        na.strings = c("", "NA", "nd"))

# Using only data from the FRRP source
# Removing observations where the response variable is missing
final_data1 <- bleaching_data %>%
  filter(!is.na(Percent_Bleaching)) %>%
  filter(Data_Source == "FRRP") %>%
  distinct(Site_ID, Sample_ID, .keep_all = TRUE)

#Removing unwanted variables
final_data2 <- final_data1 %>%
  subset(select = -c(Site_ID, Sample_ID, Data_Source, Latitude_Degrees,
                     Longitude_Degrees, Ocean_Name, Reef_ID,
                     Realm_Name, Ecoregion_Name, Country_Name,
                     State_Island_Province_Name, City_Town_Name,
                     Site_Name, Date_Day, Date_Month, Date_Year,
                     Substrate_Name, Percent_Cover,
                     Bleaching_Level, Temperature_Mean,
                     Temperature_Minimum, Temperature_Maximum,
                     Temperature_Kelvin_Standard_Deviation,
                     SSTA_Standard_Deviation, SSTA_Mean,
                     SSTA_Minimum, SSTA_Maximum, SSTA_Frequency,
                     SSTA_Frequency_Standard_Deviation, SSTA_FrequencyMax,
                     SSTA_FrequencyMean, SSTA_DHW_Standard_Deviation,
                     SSTA_DHWMax, SSTA_DHWMean, TSA_Standard_Deviation,
                     TSA_Minimum, TSA_Maximum, TSA_Mean, TSA_Frequency,
                     TSA_Frequency_Standard_Deviation, TSA_FrequencyMax,
                     TSA_FrequencyMean, TSA_DHW_Standard_Deviation,
                     TSA_DHWMax, TSA_DHWMean, Date, Site_Comments,
                     Sample_Comments, Bleaching_Comments))

# Final data set
final_data <- final_data2 %>%
  dplyr::select(
    Distance_to_Shore,
    Turbidity,
    Cyclone_Frequency,
    Depth_m,
    Windspeed,
    Temperature_Kelvin,
    ClimSST,
    SSTA,
    SSTA_DHW,
    TSA,
    TSA_DHW,
    Exposure,
    Percent_Bleaching
  )

final_data$Exposure <- as.factor(final_data$Exposure)
levels(final_data$Exposure)
```

# Data Exploration

**Below we will explore the data!**

```{r}
# Density plot of Percent_Bleaching
plot(density(final_data$Percent_Bleaching),
     main = "Density of Percent_Bleaching",
     xlab = "Y = Percent_Bleaching")
```

```{r}
# Looking at the correlations of all numeric variables
cor(final_data[,-c(12,13)])

# Correlation plot
corrplot::corrplot(corr = cor(final_data[,-c(12,13)]))
```

Below we will investigate the correlation of each variable with the response variable by using scatter plots and box plots.

```{r}
par(mfrow = c(2,3))

#First set of graphs
plot(x = final_data$Distance_to_Shore,
     y = final_data$Percent_Bleaching,
     xlab = "Distance to Shore",
     ylab = "Percent Bleaching")
plot(x = final_data$Turbidity,
     y = final_data$Percent_Bleaching,
     xlab = "Turbidity",
     ylab = "Percent Bleaching")
plot(x = final_data$Cyclone_Frequency,
     y = final_data$Percent_Bleaching,
     xlab = "Cyclone Frequency",
     ylab = "Percent Bleaching")
plot(x = final_data$Depth_m,
     y = final_data$Percent_Bleaching,
     xlab = "Depth (Meters)",
     ylab = "Percent Bleaching")
plot(x = final_data$Temperature_Kelvin,
     y = final_data$Percent_Bleaching,
     xlab = "Temperature (Kelvin)",
     ylab = "Percent Bleaching")
plot(x = final_data$ClimSST,
     y = final_data$Percent_Bleaching,
     xlab = "ClimSST",
     ylab = "Percent Bleaching")

# Second set of graphs
plot(x = final_data$SSTA,
     y = final_data$Percent_Bleaching,
     xlab = "SSTA",
     ylab = "Percent Bleaching")
plot(x = final_data$SSTA_DHW,
     y = final_data$Percent_Bleaching,
     xlab = "SSTA_DHW",
     ylab = "Percent Bleaching")
plot(x = final_data$TSA,
     y = final_data$Percent_Bleaching,
     xlab = "TSA",
     ylab = "Percent Bleaching")
plot(x = final_data$TSA_DHW,
     y = final_data$Percent_Bleaching,
     xlab = "TSA_DHW",
     ylab = "Percent Bleaching")
boxplot(Percent_Bleaching ~ Windspeed,
        data = final_data, 
        ylab = "Percent Bleaching")
boxplot(Percent_Bleaching ~ Exposure,
        data = final_data, 
        ylab = "Percent Bleaching")
```

* We will remove `TSA` because it is highly correlated with `Temperature_Kelvin`, having a correlation of 0.932.
* We will remove `TSA_DHW` because it is highly correlated with `SSTA_DHW`, having a correlation of 0.952.

```{r}
par(mfrow = c(2,3))

plot(density(final_data$Distance_to_Shore),
     main = "Density of Distance_to_Shore")
plot(density(final_data$Turbidity),
     main = "Density of Turbidity")
plot(density(final_data$Cyclone_Frequency),
     main = "Density of Cyclone_Frequency")
plot(density(final_data$Depth_m),
     main = "Density of Depth_m")
plot(density(final_data$Windspeed),
     main = "Density of Windspeed")
plot(density(final_data$Temperature_Kelvin),
     main = "Density of Temperature_Kelvin")

plot(density(final_data$ClimSST),
     main = "Density of ClimSST")
plot(density(final_data$SSTA),
     main = "Density of SSTA")
plot(density(final_data$SSTA_DHW),
     main = "Density of SSTA_DHW")
plot(density(final_data$TSA),
     main = "Density of TSA")
plot(density(final_data$TSA_DHW),
     main = "Density of TSA_DHW")
```

# Data Splitting

**Below we will split the data!**

```{r}
# Remove more unwanted variables
coral_data <- final_data %>%
  subset(select = -c(TSA, TSA_DHW))
```

```{r}
set.seed(123)

# Splitting the data into training and test sets
index <- createDataPartition(coral_data$Percent_Bleaching,
                             p = 0.75,
                             list = FALSE,
                             times = 1)
train <- coral_data[index, ]
test <- coral_data[-index, ]
```

# Data Analysis

**Below we will model the data!**

## (Section 1) Linear Regression Models

### Model 1.1: Multiple Linear Regression (MLR)

Below we will have two different models, a regression model without interaction terms and a regression model with interaction terms.

* Model 1.1a: Multiple Linear Regression (without interaction terms)
* Model 1.1b: Multiple Linear Regression (with interaction terms)

#### Model 1.1a: Multiple Linear Regression (without interaction terms)

Below is a multiple linear regression model without interaction terms.

```{r}
###################################### Model 1.1a ######################################
# Model 1.1a: Multiple Linear Regression (without interaction terms)

set.seed(123)

# Fitting the model
lm_model1 <- lm(Percent_Bleaching ~ .,
                data = train)
summary(lm_model1)
```

We can see above that the following variables are NOT significant:

* `Windspeed`
* `ClimSST`
* `Exposure`

Below we will recreate the model without these variables.

```{r}
# Fitting the model
lm_model1_final <- lm(Percent_Bleaching ~ Distance_to_Shore +
                        Turbidity +
                        Cyclone_Frequency +
                        Depth_m +
                        Temperature_Kelvin +
                        SSTA +
                        SSTA_DHW,
                      data = train)
summary(lm_model1_final)
```

```{r}
# Predict response for the test (holdout) set
pred <- predict(lm_model1_final, newdata = test)

# Compute test MSE
lm_error1 <- mean((test$Percent_Bleaching - pred)**2)
paste("Our estimated test MSE is", round(lm_error1, 2))

# Compute test RMSE
lm_rmse1 <- sqrt(lm_error1)
paste("Our estimated test RMSE is", round(lm_rmse1, 2))

# Compute R_squared
r_squared_lm_model1 <- round(cor(pred, test$Percent_Bleaching)**2, 4)
paste("Our estimated R_squared is", round(r_squared_lm_model1, 4))
```

```{r}
par(mfrow = c(2, 2))
plot(lm_model1_final)
```

#### Model 1.1b: Multiple Linear Regression (with interaction terms)

Below is a multiple linear regression model with interaction terms.

```{r}
###################################### Model 1.1b ######################################
# Model 1.1b: Multiple Linear Regression (with interaction terms)

set.seed(123)

# Fitting the model
lm_model2 <- lm(Percent_Bleaching ~ (.)^2,
                data = train)
#summary(lm_model2)

lm_model2_final <- lm(Percent_Bleaching ~ Distance_to_Shore*Temperature_Kelvin +
                        Distance_to_Shore*ClimSST +
                        Turbidity*Depth_m +
                        Cyclone_Frequency*Windspeed +
                        Depth_m*Temperature_Kelvin +
                        Temperature_Kelvin*SSTA_DHW +
                        ClimSST*SSTA_DHW +
                        SSTA*SSTA_DHW,
                      data = train)
summary(lm_model2_final)
```

```{r}
# Predict response for the test (holdout) set
pred <- predict(lm_model2_final, newdata = test)

# Compute test MSE
lm_error2 <- mean((test$Percent_Bleaching - pred)**2)
paste("Our estimated test MSE is", round(lm_error2, 2))

# Compute test RMSE
lm_rmse2 <- sqrt(lm_error2)
paste("Our estimated test RMSE is", round(lm_rmse2, 2))

# Compute R_squared
r_squared_lm_model2 <- round(cor(pred, test$Percent_Bleaching)**2, 4)
paste("Our estimated R_squared is", round(r_squared_lm_model2, 4))
```

```{r}
par(mfrow = c(2, 2))
plot(lm_model2_final)
```

### Model 1.2: Subset Selection

Note that all of the three subset selection models chose models of size 7 with the following variables:

* `Distance_to_Shore`
* `Turbidity`
* `Cyclone_Frequency`
* `Depth_m`
* `Temperature_Kelvin`
* `SSTA`
* `SSTA_DHW`

`Windspeed`, `ClimSST`, and `Exposure` were not included in any of the models.

**Note that we already fit a linear model with these 7 variables above. It is Model 1.1a.**

#### Model 1.2a: Best Subset Selection

Below we will execute best subset selection.

```{r}
###################################### Model 1.2a ######################################
# Model 1.2a: Best Subset Selection

set.seed(123)

# Fitting the model
best_subset <- regsubsets(Percent_Bleaching ~ ., 
                          data = train, 
                          nvmax = 10)

# Model summary
summary_best_subset <- summary(best_subset)

# Creating a data frame of performance metrics
metrics_best <- data.frame(model_size = seq(from = 1, to = 10, by = 1),
                           aic = summary_best_subset$cp,
                           bic = summary_best_subset$bic,
                           adjR2 = summary_best_subset$adjr2)
metrics_best

# Finding which model has the lowest Cp and BIC
c("Lowest Cp" = which.min(summary_best_subset$cp), 
  "Lowest BIC" = which.min(summary_best_subset$bic),
  "Highest Adj R2" = which.max(summary_best_subset$adjr2))
```

We will use a model of size 7 because it has a higher $R^2$ than the model of size 4.

```{r}
set.seed(123)

# Fitting the model
best_subset_final <- regsubsets(Percent_Bleaching ~ ., 
                                data = test, 
                                nvmax = 10)

# Model summary
summary_best_subset_final <- summary(best_subset_final)

# Reporting coefficients of best model
coef(best_subset_final, 7)
```

#### Model 1.2b: Forward Stepwise Selection

Below we will execute forward stepwise selection.

```{r}
###################################### Model 1.2b ######################################
# Model 1.2b: Forward Stepwise Selection

set.seed(123)

# Fitting the model
forward_subset <- regsubsets(Percent_Bleaching ~ ., 
                             data = train, 
                             nvmax = 10,
                             method = "forward")

# Model summary
summary_forward <- summary(forward_subset)

# Creating a data frame of performance metrics
metrics_forward <- data.frame(model_size = seq(from = 1, to = 10, by = 1),
                              aic = summary_forward$cp,
                              bic = summary_forward$bic,
                              adjR2 = summary_forward$adjr2)
metrics_forward

# Finding which model has the lowest Cp and BIC
c("Lowest Cp" = which.min(summary_forward$cp), 
  "Lowest BIC" = which.min(summary_forward$bic),
  "Highest Adj R2" = which.max(summary_forward$adjr2))
```

We will use a model of size 7 because it has a higher $R^2$ than the model of size 4.

```{r}
set.seed(123)

# Fitting the model
forward_final <- regsubsets(Percent_Bleaching ~ ., 
                            data = test, 
                            nvmax = 10,
                            method = "forward")

# Model summary
summary_forward_final<- summary(forward_final)

# Reporting coefficients of best model
coef(forward_final, 7)
```

#### Model 1.2c: Backward Stepwise Selection

Below we will execute backward stepwise selection.

```{r}
###################################### Model 1.2c ######################################
# Model 1.2c: Backward Stepwise Selection

set.seed(123)

# Fitting the model
backward_subset <- regsubsets(Percent_Bleaching ~ ., 
                             data = train, 
                             nvmax = 10,
                             method = "backward")

# Model summary
summary_backward <- summary(backward_subset)

# Creating a data frame of performance metrics
metrics_backward <- data.frame(model_size = seq(from = 1, to = 10, by = 1),
                               aic = summary_backward$cp,
                               bic = summary_backward$bic,
                               adjR2 = summary_backward$adjr2)
metrics_backward

# Finding which model has the lowest Cp and BIC
c("Lowest Cp" = which.min(summary_backward$cp), 
  "Lowest BIC" = which.min(summary_backward$bic),
  "Highest Adj R2" = which.max(summary_backward$adjr2))
```

We will use a model of size 7 because it has a higher $R^2$ than the model of size 4.

```{r}
set.seed(123)

# Fitting the model
backward_final <- regsubsets(Percent_Bleaching ~ ., 
                             data = test, 
                             nvmax = 10,
                             method = "backward")

# Model summary
summary_backward_final <- summary(backward_final)

# Reporting coefficients of best model
coef(backward_final, 7)
```

### Model 1.3: Ridge Regression

Below we will fit a ridge regression model with $\lambda$ chosen by cross-validation.

```{r}
####################################### Model 1.3 #######################################
# Model 1.3: Ridge Regression

set.seed(123)

X_train <- model.matrix(Percent_Bleaching ~ ., data = train)[, -1]
X_test <- model.matrix(Percent_Bleaching ~ ., data = test)[, -1]

Y_train <- train$Percent_Bleaching
Y_test <- test$Percent_Bleaching

# Grid of values for lambda
grid <- 10^seq(-2, 8, length = 100)

# Fitting the model
ridge_fit <- cv.glmnet(x = X_train,
                       y = Y_train,
                       alpha = 0,
                       lambda = grid)
plot(ridge_fit)

# Optimal lambda value
lambda_min <- ridge_fit$lambda.min
lambda_1se <- ridge_fit$lambda.1se

# Prediction and finding test MSE
ridge_pred <- predict(ridge_fit, s = lambda_1se, newx = X_test)
ridge_error <- mean( (test$Percent_Bleaching - ridge_pred)^2 )
paste("Our estimated test MSE is", round(ridge_error, 2))

# Compute test RMSE
ridge_rmse <- sqrt(ridge_error)
paste("Our estimated test RMSE is", round(ridge_rmse, 2))
```

```{r}
# Reporting the coefficients
predict(ridge_fit, s = lambda_1se, type = "coefficients")[1:11, ]
```

### Model 1.4: LASSO Regression

Below we will fit a LASSO regression model with $\lambda$ chosen by cross-validation.

```{r}
####################################### Model 1.4 #######################################
# Model 1.4: LASSO Regression

set.seed(123)

X_train <- model.matrix(Percent_Bleaching ~ ., data = train)[, -1]
X_test <- model.matrix(Percent_Bleaching ~ ., data = test)[, -1]

Y_train <- train$Percent_Bleaching
Y_test <- test$Percent_Bleaching

# Grid of values for lambda
grid <- 10^seq(-2, 8, length = 100)

# Fitting the model
lasso_fit <- cv.glmnet(x = X_train,
                       y = Y_train,
                       alpha = 1,
                       lambda = grid)
plot(lasso_fit)

# Optimal lambda value
lambda_min <- lasso_fit$lambda.min
lambda_1se <- lasso_fit$lambda.1se

# Prediction and finding test MSE
lasso_pred <- predict(lasso_fit, s = lambda_1se, newx = X_test)
lasso_error <- mean( (test$Percent_Bleaching - lasso_pred)^2 )
paste("Our estimated test MSE is", round(lasso_error, 2))

# Compute test RMSE
lasso_rmse <- sqrt(lasso_error)
paste("Our estimated test RMSE is", round(lasso_rmse, 2))
```

```{r}
# Reporting the coefficients
predict(lasso_fit, s = lambda_1se, type = "coefficients")[1:11, ]
```

The only non-zero coefficients are for the following variables:

* `Temperature_Kelvin`
* `SSTA_DHW`

### Model 1.5: Principal Components Regression (PCR)

Below we will fit a principal components regression model with $M$ chosen by cross-validation.

```{r}
####################################### Model 1.5 #######################################
# Model 1.5: Principal Components Regression (PCR)

set.seed(123)

# Fitting the model
pcr_fit <- pcr(Percent_Bleaching ~ .,
               data = train,
               center = TRUE,
               scale=TRUE,
               validation = "CV")
summary(pcr_fit)
```

We can see above (and as is the case for PCR) we do not have 100% of our variance explained until we get
to all $M = p = 10$ components, which is also where the lowest cross-validation error occurs. To combat
including all 10 components, we can say that we would like to have at least 85% of our variance explained
by our model, and therefore we would chose a model with $M = 7$ components!

```{r}
# Prediction and finding test MSE (using 7 components)
pcr_pred <- predict(pcr_fit, test, ncomp = 7)
pcr_error <- mean( (test$Percent_Bleaching - pcr_pred)**2 )
paste("Our estimated test MSE is", round(pcr_error, 2))

# Compute test RMSE
pcr_rmse <- sqrt(pcr_error)
paste("Our estimated test RMSE is", round(pcr_rmse, 2))
```

### Model 1.6: Partial Least Squares Regression (PLS)

Below we will fit a partial least squares regression model with $M$ chosen by cross-validation.

```{r}
####################################### Model 1.6 #######################################
# Model 1.6: Partial Least Squares Regression (PLS)

set.seed(123)

# Fitting the model
pls_fit <- plsr(Percent_Bleaching ~ .,
                data = train,
                center = TRUE,
                scale=TRUE,
                validation = "CV")
summary(pls_fit)
```

We can see above that once we get to $M = 5$ components, we reach the lowest cross-validation error. However, we do not have at least 85% of our variance explained until we get to $M = 8$ components. Therefore, we will choose a model with $M = 8$ components!

```{r}
# Prediction and finding test MSE (using 8 components)
pls_pred <- predict(pls_fit, test, ncomp = 8)
pls_error <- mean( (test$Percent_Bleaching - pls_pred)**2 )
paste("Our estimated test MSE is", round(pls_error, 2))

# Compute test RMSE
pls_rmse <- sqrt(pls_error)
paste("Our estimated test RMSE is", round(pls_rmse, 2))
```

## (Section 2) Nonparametric Regression Models

### Model 2.1: K-Nearest Neighbors (KNN)

Below we will fit a KNN model with $K$ chosen by cross-validation.

```{r}
####################################### Model 2.1 #######################################
# Model 2.1: KNN Regression (KNN)

set.seed(123)

# Setting K grid
kgrid <- expand.grid(k = c(1:100))

# Training control parameters
cv <- trainControl(method = "cv",
                   number = 5)

# Fitting the model
knn_fit <- train(Percent_Bleaching ~ .,
                 data = train,
                 method = "knn",
                 tuneGrid = kgrid,
                 trControl = cv)

# Plotting knn_fit
plot(knn_fit)

# Finding optimal K
k_opt <- knn_fit$bestTune$k
paste("Our optimal K value is", k_opt)

# Refitting with optimal K
knn_fit <- train(Percent_Bleaching ~ .,
                   data = train,
                   method = "knn",
                   tuneGrid = expand.grid(k = k_opt),
                   trControl = trainControl(method = "none"))
```

```{r}
# Predict response for the test (holdout) set
knn_pred <- predict(knn_fit, newdata = test)

# Compute test MSE
knn_error <- mean((test$Percent_Bleaching - knn_pred)**2)
paste("Our estimated test MSE is", round(knn_error, 2))

# Compute test RMSE
knn_rmse <- sqrt(knn_error)
paste("Our estimated test RMSE is", round(knn_rmse, 2))
```

### Model 2.2: Multivariate Adaptive Regression Splines (MARS)

Below we will fit a multivariate adaptive regression splines model.

```{r}
####################################### Model 2.2 #######################################
# Model 2.2: Multivariate Adaptive Regression Splines (MARS)

set.seed(123)

# Create tuning grid
grid <- expand.grid(degree = 1:3,
                    nprune = seq(2, 50, length.out = 10) %>%
                      floor())

# Fitting the model
mars_fit <- train(Percent_Bleaching ~ .,
                  data = train,
                  method = "earth",
                  metric = "RMSE",
                  trControl = trainControl(method = "cv", number = 10),
                  tuneGrid = grid)
ggplot(mars_fit)

# Displaying model with lowest training RMSE
mars_fit$results %>%
  filter(nprune == mars_fit$bestTune$nprune,
         degree == mars_fit$bestTune$degree)
```

```{r}
# Predict response for the test (holdout) set
mars_pred <- predict(mars_fit, newdata = test)

# Compute test MSE
mars_error <- mean((test$Percent_Bleaching - mars_pred)**2)
paste("Our estimated test MSE is", round(mars_error, 2))

# Compute test RMSE
mars_rmse <- sqrt(mars_error)
paste("Our estimated test RMSE is", round(mars_rmse, 2))
```

### Model 2.3: Generalized Additive Model (GAM)

Below we will fit a generalized additive model.

```{r}
####################################### Model 2.3 #######################################
# Model 2.3: Generalized Additive Model (GAM)

set.seed(123)

require(gam)
require(splines)

# Fitting the model
gam_fit <- gam(Percent_Bleaching ~ Exposure
                 + s(Distance_to_Shore)
                 + s(Turbidity)
                 + s(Cyclone_Frequency)
                 + s(Depth_m)
                 + s(Windspeed)
                 + s(Temperature_Kelvin)
                 + s(ClimSST)
                 + s(SSTA)
                 + s(SSTA_DHW),
               data = train)
summary(gam_fit)

par(mfrow = c(2,3))
plot(gam_fit, se = TRUE, col = "red")
```

We have very strong nonlinear relationships between `Percent_Bleaching` and both `ClimSST` and `SSTA_DHW`. We have a moderately strong nonlinear relationship between `Percent_Bleaching` and `Temperature_Kelvin`.

```{r}
# Predict response for the test (holdout) set
gam_pred <- predict(gam_fit, newdata = test)

# Compute test MSE
gam_error <- mean((test$Percent_Bleaching - gam_pred)**2)
paste("Our estimated test MSE is", round(gam_error, 2))

# Compute test RMSE
gam_rmse <- sqrt(gam_error)
paste("Our estimated test RMSE is", round(gam_rmse, 2))
```

### Model 2.4: Local Regression (LR)

Below we will fit a local regression model.

```{r}
####################################### Model 2.4 #######################################
# Model 2.4: Local Regression (LR)

set.seed(123)

# Training control parameters and tuning grid
cv <- trainControl(method = "cv", number = 10)
grid <- expand.grid(span = seq(0.5, 0.9, len = 5), degree = 1)

# Fitting the model
local_model <- train(Percent_Bleaching ~ .,
                     data = train,
                     method = "gamLoess",
                     trControl = cv,
                     tuneGrid = grid)
local_model
```

We can see above that the optimal value for span is 0.5.

```{r}
set.seed(123)

# Fitting the model
local_model2 <- train(Percent_Bleaching ~ .,
                     data = train,
                     method = "gamLoess",
                     trControl = trainControl(method = "none"),
                     tuneGrid = expand.grid(span = 0.5, degree = 1))
```

```{r}
# Predict response for the test (holdout) set
local_pred <- predict(local_model2, newdata = test)

# Compute test MSE
local_error <- mean((test$Percent_Bleaching - local_pred)**2)
paste("Our estimated test MSE is", round(local_error, 2))

# Compute test RMSE
local_rmse <- sqrt(local_error)
paste("Our estimated test RMSE is", round(local_rmse, 2))
```

### Model 2.5: Support Vector Regression (SVR)

Below we will fit a support vector regression model.

```{r}
####################################### Model 2.5 #######################################
# Model 2.5: Support Vector Machine (SVM)

set.seed(123)

svr_fit <- svm(Percent_Bleaching ~ .,
               data = train,
               type = "eps-regression",
               kernel = "radial",
               gamma = 1,
               cost = 1,
               epsilon = 0.1)
summary(svr_fit)

```

```{r}
# Predict response for the test (holdout) set
svr_pred <- predict(svr_fit, newdata = test)

# Compute test MSE
svr_error <- mean((test$Percent_Bleaching - svr_pred)**2)
paste("Our estimated test MSE is", round(svr_error, 2))

# Compute test RMSE
svr_rmse <- sqrt(svr_error)
paste("Our estimated test RMSE is", round(svr_rmse, 2))
```

## (Section 3) Regression Tree Models

### Model 3.1: Standard Regression Tree

Below we will fit a standard regression tree model.

```{r}
####################################### Model 3.1 #######################################
# Model 3.1: Standard Regression Tree

set.seed(123)

# Building the initial tree
tree_fit <- rpart(Percent_Bleaching ~ .,
                  data = train,
                  control = rpart.control(cp=.0001))
#printcp(tree_fit)

# Identify the best cp value to use
best_cp <- tree_fit$cptable[which.min(tree_fit$cptable[,"xerror"]),"CP"]

# Pruned tree based on the best cp value
pruned_tree <- prune(tree_fit, cp = best_cp)

#jpeg(file="regressiontree.jpeg")

# Plot the pruned tree
tree_plot <- prp(pruned_tree,
                 faclen = 0, # Use full names for factor labels
                 extra = 1, # Display number of obs for each terminal node
                 roundint = F, # Don't round to integers in output
                 digits = 5) # Display 5 decimal places in output
#dev.off()
```

```{r}
# Predict response for the test (holdout) set
tree_pred <- predict(tree_fit, newdata = test)

# Compute test MSE
tree_error <- mean((test$Percent_Bleaching - tree_pred)**2)
paste("Our estimated test MSE is", round(tree_error, 2))

# Compute test RMSE
tree_rmse <- sqrt(tree_error)
paste("Our estimated test RMSE is", round(tree_rmse, 2))
```

### Model 3.2: Random Forest

Below we will fit a random forest model.

```{r}
####################################### Model 3.2 #######################################
# Model 3.2: Random Forest

set.seed(123)

# Fitting the model
rf_fit <- randomForest(formula = Percent_Bleaching ~ .,
                       data = train)

# Number of trees that produced lowest test MSE
which.min(rf_fit$mse)

# Plotting the error vs number of trees
plot(rf_fit, main = "Random Forest Model")
points(which.min(rf_fit$mse), rf_fit$mse[which.min(rf_fit$mse)],
       col = "red", cex = 2, pch = 20)

# Variable importance plot
varImpPlot(rf_fit, main = "Random Forest Model") 
```

Below we will tune the model.

```{r}
set.seed(123)

rf_tuned <- tuneRF(x = train[,-1],
                   y = train$Percent_Bleaching,
                   ntreeTry = 500,
                   mtryStart = 4,
                   stepFactor = 1.5,
                   improve = 0.01)
rf_tuned
```

The lowest OOB error is achieved by using 10 randomly selected predictors at each split when building the trees.

```{r}
# Predict response for the test (holdout) set
rf_pred <- predict(rf_fit, newdata = test)

# Compute test MSE
rf_error <- mean((test$Percent_Bleaching - rf_pred)**2)
paste("Our estimated test MSE is", round(rf_error, 2))

# Compute test RMSE
rf_rmse <- sqrt(rf_error)
paste("Our estimated test RMSE is", round(rf_rmse, 2))
```

### Model 3.3: Bagging

Below we will fit a bagged tree model.

```{r}
####################################### Model 3.3 #######################################
# Model 3.3: Bagging

set.seed(123)

bag_fit <- bagging(formula = Percent_Bleaching ~ .,
                   data = train,
                   nbagg = 150,
                   coob = TRUE,
                   control = rpart.control(minsplit = 2, cp = 0))
bag_fit
```

Meaning of certain arguments:

* `nbagg`: Number of bootstrap samples.
* `coob`: Set to `TRUE` in order to obtain the estimated OOB error.
* `minsplit`: This tells the model to only require 2 observations in a node to split.
* `cp`: This is the complexity parameter. By setting it to 0, we don’t require the model to be able to improve the overall fit by any amount in order to perform a split.

Essentially, these last two arguments allow the individual trees to grow extremely deep, which leads to trees with high variance, but low bias. Then when we apply bagging we’re able to reduce the variance of the final model while keeping the bias low.

```{r}
set.seed(123)

# Calculate variable importance
VI <- data.frame(var = names(coral_data[,-1]), imp = varImp(bag_fit))

# Sort variable importance descending
VI_plot <- VI[order(VI$Overall, decreasing=TRUE),]

par(mar=c(4,10,4,4))

# Visualize variable importance with horizontal bar plot
barplot(VI_plot$Overall,
        names.arg = rownames(VI_plot),
        horiz = TRUE,
        col = "steelblue",
        xlab = "Variable Importance",
        las = 2)
```

```{r}
# Predict response for the test (holdout) set
bag_pred <- predict(bag_fit, newdata = test)

# Compute test MSE
bag_error <- mean((test$Percent_Bleaching - bag_pred)**2)
paste("Our estimated test MSE is", round(bag_error, 2))

# Compute test RMSE
bag_rmse <- sqrt(bag_error)
paste("Our estimated test RMSE is", round(bag_rmse, 2))
```

### Model 3.4: Boosting

Below we will fit a boosted tree model.

```{r}
####################################### Model 3.4 #######################################
# Model 3.4: Boosting

set.seed(123)

boost_fit <- gbm(Percent_Bleaching ~ .,
                 data = train,
                 distribution = "gaussian",
                 n.trees = 5000,
                 interaction.depth = 4)
summary(boost_fit)
```

```{r}
# Predict response for the test (holdout) set
boost_pred <- predict(boost_fit, newdata = test)

# Compute test MSE
boost_error <- mean((test$Percent_Bleaching - boost_pred)**2)
paste("Our estimated test MSE is", round(boost_error, 2))

# Compute test RMSE
boost_rmse <- sqrt(boost_error)
paste("Our estimated test RMSE is", round(boost_rmse, 2))
```

### Model 3.5: Bayesian Additive Regression Tree (BART)

Below we will fit a Bayesian additive regression tree model.

```{r}
####################################### Model 3.5 #######################################
# Model 3.5: Bayesian Additive Regression Tree (BART)

set.seed(123)

x <- train[,1:10]
y <- ("y" = train$Percent_Bleaching)

# Fitting the model
bart_fit <- bartMachine(X = x, y = y,
                        num_trees = 100)
summary(bart_fit)
```

```{r}
# Predict response for the test (holdout) set
bart_pred <- bart_machine_get_posterior(bart_fit, new_data = test[,1:10])

# Compute test MSE
bart_error <- mean((test$Percent_Bleaching - bart_pred$y_hat)**2)
paste("Our estimated test MSE is", round(bart_error, 2))

# Compute test RMSE
bart_rmse <- sqrt(bart_error)
paste("Our estimated test RMSE is", round(bart_rmse, 2))
```

## (Section 4) Stacking Models

### Model 4.1: GAM / Random Forest / GBM (Boosting)

Below we will fit a stacked model using GAM, Random Forest, and GBM (Boosting).

```{r}
####################################### Model 4.1 #######################################
# Model 4.1: GAM / Random Forest / Boosted Tree

set.seed(123)

# Get response and feature names
Y <- "Percent_Bleaching"
X <- setdiff(names(train), Y)

# Init h2o
h2o.init()

# Convert train/test sets to h2o format
coral_train <- as.h2o(train)
coral_test <- as.h2o(test)

nfolds <- 5
seed <- 123

# GAM
gam_model <- h2o.gam(x = X,
                     y = Y,
                     training_frame = coral_train,
                     nfolds = nfolds,
                     seed = seed,
                     keep_cross_validation_predictions = TRUE,
                     fold_assignment = "Modulo",
                     family = "gaussian",
                     gam_columns = c("Distance_to_Shore",
                                     "Turbidity",
                                     "Cyclone_Frequency",
                                     "Depth_m",
                                     "Temperature_Kelvin",
                                     "ClimSST",
                                     "SSTA"))
# Random forest
rf_model <- h2o.randomForest(x = X,
                             y = Y,
                             training_frame = coral_train,
                             nfolds = nfolds,
                             seed = seed,
                             keep_cross_validation_predictions = TRUE,
                             fold_assignment = "Modulo",
                             ntrees = 1000,
                             mtries = 5,
                             max_depth = 30,
                             sample_rate = 0.8,
                             stopping_rounds = 50,
                             stopping_metric = "RMSE",
                             stopping_tolerance = 0)
# GBM (Boosting)
gbm_model <- h2o.gbm(x = X,
                     y = Y,
                     training_frame = coral_train,
                     nfolds = nfolds,
                     seed = seed,
                     keep_cross_validation_predictions = TRUE,
                     fold_assignment = "Modulo",
                     ntrees = 5000,
                     learn_rate = 0.01,
                     max_depth = 7,
                     min_rows = 5,
                     sample_rate = 0.8,
                     stopping_rounds = 50,
                     stopping_metric = "RMSE",
                     stopping_tolerance = 0)

coral_ensemble <- h2o.stackedEnsemble(x = X,
                                      y = Y,
                                      training_frame = coral_train,
                                      model_id = "coral_ensemble",
                                      base_models = list(gam_model,
                                                         rf_model,
                                                         gbm_model),
                                      metalearner_algorithm = "AUTO")
```

```{r}
# Function to extract RMSE
get_rmse <- function(model){
  perf <- h2o.performance(model, newdata = coral_test)
  return(perf@metrics$RMSE)
}

# Function to extract MSE
get_mse <- function(model){
  perf <- h2o.performance(model, newdata = coral_test)
  return( (perf@metrics$RMSE)**2 )
}

# Model list
mod_list <- list(GAM = gam_model, 
                 RF = rf_model,
                 GBM = gbm_model,
                 STACKED = coral_ensemble)

# Apply function to model list
ensemble_rmse <- purrr::map_dbl(mod_list, get_rmse)
ensemble_error <- purrr::map_dbl(mod_list, get_mse)
```

```{r}
# Data frame of predictions
pred <- data.frame(
GAM_pred = as.vector(h2o.getFrame(gam_model@model$cross_validation_holdout_predictions_frame_id$name)),
RF_pred = as.vector(h2o.getFrame(rf_model@model$cross_validation_holdout_predictions_frame_id$name)),
GBM_pred = as.vector(h2o.getFrame(gbm_model@model$cross_validation_holdout_predictions_frame_id$name)))

# Correlation of the predictions
cor(pred)

# Plotting the correlation of the predictions
plot(pred)
```

## Summaries

### Linear Regression Models

```{r}
# Combine the error values into a vector
lm_errors <- c(lm_error1, lm_error2)
names(lm_errors) <- c("MLR Model 1.1a \n(w/o interaction terms)",
                      "MLR Model 1.1b \n(with interaction terms)")

# Creating a bar plot of the errors
bar <- barplot(lm_errors,
               cex.names = 0.9,
               xlab = "Linear Model",
               ylab = "Test MSE",
               main = "Test Errors for Different Multiple Linear Regression Models")
text(x = bar, y = lm_errors, label = round(lm_errors,2), 
     pos = 1, cex = 0.8, col = "red")
```

```{r}
# Combine the error values into a vector
errors <- c(lm_error1, lm_error2, ridge_error,
            lasso_error, pcr_error, pls_error)
names(errors) <- c("MLR \n(1.1a)",
                   "MLR \n(1.1b)",
                   "Ridge \n(1.3)",
                   "Lasso \n(1.4)",
                   "PCR \n(1.5)",
                   "PLS \n(1.6)")

# Creating a bar plot of the errors
bar <- barplot(errors,
               cex.names = 0.9,
               xlab = "Regression Method",
               ylab = "Test MSE",
               main = "Test Errors for Different Linear Regression Methods")
text(x = bar, y = errors, label = round(errors,2), 
     pos = 1, cex = 0.8, col = "red")
```

```{r}
# MSE and RMSE values for each linear model
MSE <- c(lm_error1, lm_error2, ridge_error,
                lasso_error, pcr_error, pls_error)
RMSE <- c(lm_rmse1, lm_rmse2, ridge_rmse,
                 lasso_rmse, pcr_rmse, pls_rmse)

# Combining the two vectors and adding row names
results <- data.frame("MSE" = MSE,
                      "RMSE" = RMSE)
rownames(results) <- c("MLR (w/o interaction)",
                       "MLR (with interaction)",
                       "Ridge", "LASSO",
                       "PCR", "PLS")
results
kable(results)
```

### Nonparametric Regression Models

```{r}
# Combine the error values into a vector
errors <- c(knn_error, mars_error,
            gam_error, local_error,
            svr_error)
names(errors) <- c("KNN \n(2.1)",
                   "MARS \n(2.2)",
                   "GAM \n(2.3)",
                   "LR \n(2.4)",
                   "SVR \n(2.5)")

# Creating a bar plot of the errors
bar <- barplot(errors,
               cex.names = 0.9,
               xlab = "Nonparametric Regression Method",
               ylab = "Test MSE",
               main = "Test Errors for Different Nonparametric Regression Methods")
text(x = bar, y = errors, label = round(errors,2), 
     pos = 1, cex = 0.8, col = "red")
```

```{r}
# MSE and RMSE values for each nonparametric model
MSE <- c(knn_error, mars_error,
         gam_error, local_error,
         svr_error)
RMSE <- c(knn_rmse, mars_rmse,
          gam_rmse, local_rmse,
          svr_rmse)

# Combining the two vectors and adding row names
results <- data.frame("MSE" = MSE,
                      "RMSE" = RMSE)
rownames(results) <- c("KNN", "MARS",
                       "GAM", "LR",
                       "SVR")

results
kable(results)
```

### Regression Tree Models

```{r}
# Combine the error values into a vector
tree_errors <- c(tree_error, rf_error,
                 bag_error, boost_error,
                 bart_error)
names(tree_errors) <- c("Regression \nTree (3.1)",
                        "Random \nForest (3.2)",
                        "Bagging \n(3.3)",
                        "Boosting \n(3.4)",
                        "BART \n(3.5)")

# Creating a bar plot of the errors
bar <- barplot(tree_errors,
               cex.names = 0.9,
               xlab = "Regression Tree Method",
               ylab = "Test MSE",
               main = "Test Errors for Different Regression Tree Methods")
text(x = bar, y = tree_errors, label = round(tree_errors,2), 
     pos = 1, cex = 0.8, col = "red")
```

```{r}
# MSE and RMSE values for each regression tree model
tree_MSE <- c(tree_error, rf_error,
              bag_error, boost_error,
              bart_error)
tree_RMSE <- c(tree_rmse, rf_rmse,
               bag_rmse, boost_rmse,
               bart_rmse)

# Combining the two vectors and adding row names
tree_results <- data.frame("MSE" = tree_MSE,
                           "RMSE" = tree_RMSE)
rownames(tree_results) <- c("Regression Tree",
                            "Random Forest",
                            "Bagging",
                            "Boosting",
                            "BART")

tree_results
kable(tree_results)
```

### Stacked Model

```{r}
# Creating a bar plot of the errors
bar <- barplot(ensemble_error,
               cex.names = 0.9,
               xlab = "Regression Method",
               ylab = "Test MSE",
               main = "Test Errors for Different Regression Methods & Stacking")
text(x = bar, y = ensemble_error, label = round(ensemble_error,2), 
     pos = 1, cex = 0.8, col = "red")
```

```{r}
# Combining vectors into a data frame
results <- data.frame("MSE" = ensemble_error,
                      "RMSE" = ensemble_rmse)
results
kable(results)
```


